---
title: "Conformational Signature Analysis"
author: 
  - name: Oliver Crook
    affiliation:
    - University of Oxford
    email: ococrook@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('RexMS')`"
vignette: >
  %\VignetteIndexEntry{Conformational Signature Analysis with RexMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

# Introduction

This vignette explains how to perform conformational signature analysis (CSA).
The essential idea is when you have an experimental design in which you have numerous states (such as ligand binding, mutants, antibodies) and in each state you performed
HDX-MS and potentially measured a downstream outcome. You can use conformational
signature analysis to identify the regions of the protein that are most
likely to be associated for the observed outcome. This
can then inform design choices. There are three options 
currently implemented:

1) **Unsupervised Conformational Signature Analysis** - This is the most basic
form of the analysis. It will identify the regions of the protein that are most variable
across the states. This is useful for identifying regions that are associated
with the variation in the data. By performing the analysis
in this way you will identify the states that are most similar
and different from each other.

2) **Supervised Conformational Signature Analysis with discrete outcomes** - This
is a more advanced form of the analysis using a discrete outcome (such as "monovalent" or "antagonist").
It will identify the regions of the protein that are most variable across
the states and are also associated with the outcome. This is useful
for identifying regions that are associated with the outcome. This is
done by dimensionality reduction that accounts for the outcome.

3) **Supervised Conformational Signature Analysis with continuous outcomes** - 
This is a more advanced form of the analysis using a continuous outcome
such as binding affinity. It will identify the regions of
the protein that are most variable across the states and are also associated
with the outcome. This is useful for identifying regions that are associated
with the outcome. This is done by dimensionality reduction that accounts for 
the continuous outcome. 

The analysis currently supports missing values in the outcomes
but not the HDX-MS data.

The trickest part of the analysis is setting up the data correctly. You can do
this in a excel file and load as csv but we will demonstrate using R manually below. Essentially you want to create a dataset where the rownames are the different states and columns indicate the labels associated with the states. 


# Getting started

First, we will load the library and the data. The data is a list of dataframes
where each dataframe is a state. Each dataframe should have the following columns

1) Sequence - The peptide sequence
2) Start - The start of the peptide
3) End - The end of the peptide
4) Exposure - The timepoint of the HDX-MS experiment
5) Uptake - The deuteration of the peptide
6) State - The state of the HDX-MS experiment
7) MaxUptake - The maximum deuteration of the peptide
8) replicate - The replicate of the HDX-MS experiment

This is as in the single protein analysis. We suggest performing that
analysis carefully, first, before proceeding to the conformational signature analysis.
We recommend storing the reference or apo state in a separate dataframe.
The analysis expects consistency between the experiments (e.g. replicates
and number of timepoints, region of the protein) but you do not need to measure the exact same peptides in each experiment. You will also need to name
the states you are interested in and store them in a vector.

```{r,}
library(RexMS)

# for conformation signature analysis organise data into a list

data("LXRalpha_compounds") # data for different compounds as a list
data("LXRalpha_processed") # data for the reference state

# usual important terms
numTimepoints <- length(unique(LXRalpha_compounds[[1]]$Exposure))
Timepoints <- unique(LXRalpha_compounds[[1]]$Exposure)
numPeptides <- length(unique(LXRalpha_compounds[[1]]$Sequence))
R <- max(LXRalpha_compounds[[1]]$End)

# names of compounds
states <- names(LXRalpha_compounds)

# let look at states
states
```
Our analysis has 18 compounds and a reference state. We will use the reference state
as the reference for the conformational signature analysis. Whichever route you
follow (e.g. supervised or unsupervised) the initial steps are identical. We will start with the list in an
`lapply` function. To each element in this list (which is an entire HDX dataset)
we apply the `rex` function. This function will perform the usual rex model.
There are only 3 timepoints in this analysis so we will set the `tCoef` to
`c(0, 1, 1)`. We will also set the seed to 1 for reproducibility. We will
use the `SerialParam` for the `BPPARAM` argument. This is because the analysis
is not too computationally intensive but if you run
for many iterations and chains we suggest using parrallel processing. We will store the results in a list.
For brevity we will run the analysis for 100 iterations but we recommend much 
more (typically 5000) for final analysis. The following function takes a while 
to run so we will not run them here but pre-load the results from the package
the code is placed for completeness.

```{r, eval=FALSE}
set.seed(1)
rex_lxr <- lapply(LXRalpha_compounds, function(xx) {
                
                rex(HdxData = DataFrame(xx),
                        numIter = 100,
                        R = R, 
                        numtimepoints = numTimepoints,
                        timepoints = Timepoints,
                        density = "laplace",
                        seed = 1L,
                        tCoef = c(0, rep(1, 2)),
                        BPPARAM = SerialParam())})

```

The next step is to apply `RexProcess` to the output of the `rex` function. 
Remember there are 18 compounds so we will use `lapply` again. We will set the
range to 50:100 - removing the first 50 as burnin, though we would typically recommend you run about 5000 iterations. This is mcmc iteration we are interested in. We will store the
results in a list. Again, we do not actually run the code here but pre-load the
results from the package.

```{r, eval=FALSE}

rex_lxr <- lapply(seq_along(rex_lxr), 
                            function(i)
                              RexProcess(HdxData = DataFrame(LXRalpha_compounds[[i]]),
                                          params = rex_lxr[[i]],
                                         range = 50:100))
```

We will now load the results from the package. The results are stored in the
`rex_lxr` object.

```{r}
data("rex_lxr")
```

In the next step we would perform the same differential analysis as in the
differential analysis tutorial but 18 times because there are 18 compounds. We will
use the same reference state for each compound. We will store the results in
a list. Again, we do not actually run the code here but pre-load the results
from the package. Remember `params` are the various states and the
`HdxData` is the reference state HDX data. The `params` is the output of the
`rex` function, which has been processed. The `whichChain` is the chain of the
analysis you are interested in using to generate the results. We will use chain 1,
but recommend assessing the quality of multiple chains.

```{r, eval=FALSE}
out_lxr_compound_proccessed <- lapply(rex_lxr, function(xx)
                                        processDifferential(HdxData =  DataFrame(LXRalpha_processed),
                                                            params = xx,
                                                            whichChain = 1))
```

We will now load the results from the package. The results are stored in the
`out_lxr_compound_proccessed` object.

```{r}
data("out_lxr_compound_proccessed")
```


# Unsupervised Conformational Signature Analysis

To perform the unsupervised conformational signature analysis, we will use the
`unsupervisedCSA` function. This function will take the output of the
`processDifferential` function. Whilst for unsupervised analysis you do not need
to specify the outcome, we will generate the labels results anyway to highlight
the details of setting up that analysis.

For our protein there are 18 compounds and a reference state. Each of the compounds
induce one outcome in two scenarios or the outcome is unknown (which is allowed in
our analysis). We will generate the labels for the outcomes. We will store
the results in a dataframe with the states as the rownames. For proteins where
the outcome is unknown, we will set the outcome to "Unknown". For the two scenarios,
the first is ABCA1 induction and the second is lipogencity. We will set the
outcome to "low" or "high" for ABCA1 induction. We will set the outcome to
"Non-Lipogenic" or "Lipogenic" for lipogenicity. Of course, you may adapt these
outcomes to your own analysis. We will then convert the outcomes to factors
with the levels in the order we are interested in. You can of course
load this dataframe from something that was made in excel and covert to a
dataframe.

```{r,}
# First set up the dataframe with unknowns and correct rownames
labels <- data.frame(ABCA1 = rep("Unknown", length(states)),
                     lipogenic = rep("Unknown", length(states)))
rownames(labels) <- states

# Now set the outcomes with ABCA1 induction
labels$ABCA1[rownames(labels) %in% c("LXR.623", "AZ9", "AZ8", "AZ5")] <- "low"
labels$ABCA1[rownames(labels) %in% c("Az1", "AZ2", "AZ3", "AZ4", "AZ6",
                                     "AZ7", "AZ876", "T0.901317", "WAY.254011",
                                     "F1", "GW3965", "BMS.852927")] <- "high"

# Now set the outcomes with lipogenicity
labels$lipogenic[rownames(labels) %in% c("AZ6", "AZ7", "AZ9",
                                         "AZ8", "GW3965", "BMS.852927",
                                                 "LXR.623")] <- "Non-Lipogenic"

labels$lipogenic[rownames(labels) %in% c("AZ876", "AZ1",
                                         "T0.901317", "F1", "WAY.254011")] <- "Lipogenic"

# Now convert to factors with the correct levels
labels$ABCA1 <- factor(labels$ABCA1,
                       levels = c("low", "high", "Unknown"))
labels$lipogenic <- factor(labels$lipogenic,
                           levels = c("Non-Lipogenic", "Lipogenic", "Unknown"))
```

We will now run the unsupervised conformational signature analysis. We will
set the `quantity` to "TRE" as this is the quantity we use to generate the signature.
We will set the `states` to the states we are interested in. We will set the
`whichTimepoint` to 600. This is the timepoint used to generate the conformational
signature. We will store the results in the `ucsa` object. Since this is the 
unsupervised analysis, we do not need to specify the `labels` argument.


```{r,}
ucsa <- UnsupervisedCSA(RexDifferentialList = out_lxr_compound_proccessed,
                        quantity = "TRE",
                        states = states,
                        whichTimepoint = 600)
```

We will now look at the results. The results are stored in a list with the
first entry `pca_states` being the Principal Component Analysis of the `RexMS` results.
The rotation matrix can be used as the conformational signature - providing the modes of conformational variation in the data. This contains
all the results from a PCA analysis. The second entry `states_wide` is a dataframe
with the states as the columns and the residue as the rows. The values are the
TRE values that were used to generate the conformational signature.

You can perform any downstream analysis you would like with the results. For
example, you could plot the PCA results. We will plot the first two components
of the PCA analysis. The labels are not used unless you pass them to the 
`pca_params` argument even if you supply a `labels` argument.

Note that the results look strange because we have not run the algorithm for 
enough iterations.


```{r,}
library(ggfortify)
# doesn't use the labels even though they are passed as an argument
plotUCSA(pca_states = ucsa$pca_states,
         states_wide = ucsa$states_wide,
         labels = labels)

# uses labels to colour the points
plotUCSA(pca_states = ucsa$pca_states,
         states_wide = ucsa$states_wide,
         labels = labels,
         pca_params = list(x = 1, # pca component 1 on x-axis
                           y = 2, # pca component 2 on y-axis
                           whichlabel = "ABCA1")) # colour by ABCA1


```

# Supervised Conformational Signature Analysis with discrete outcomes

Until now, we have not used the labels as part of the analysis - only to interpret
the results of the unsupervised analysis. Supervised conformational signature
analysis is more advanced and uses the labels to generate the conformational
signature. The idea is to identify the regions of the protein that are most
variable across the states and are also associated with the outcome.

To perform the supervised conformational signature analysis with discrete outcomes,
we will use the `supervisedCSA` function. This function will take the output of the
`processDifferential` function. We will generate the labels for the outcomes as
before. We will store the results in a dataframe with the states as the rownames.
For proteins where the outcome is unknown, we will set the outcome to "Unknown".
For the two scenarios, the first is ABCA1 induction and the second is lipogencity.
We will set the outcome to "low" or "high" for ABCA1 induction. We will set the
outcome to "Non-Lipogenic" or "Lipogenic" for lipogenicity. Of course, you may adapt
these outcomes to your own analysis. We will then convert the outcomes to factors
with the levels in the order we are interested in. You can of course load this
dataframe from something that was made in excel and covert to a dataframe.

We will now run the supervised conformational signature analysis with discrete
outcomes. We will set the `quantity` to "TRE" as this is the quantity we use to
generate the signature. We will set the `states` to the states we are interested in.
We will set the `whichTimepoint` to 600. This is the timepoint used to generate the
conformational signature. We will store the results in the `scsa` object. We
will set the `whichlabel` to "ABCA1" as this is the outcome we are interested in.
We will also set `whichlabel` to "lipogenic" to generate the conformational signature
for lipogenicity. Since this is the supervised analysis we need to specify the
`labels` argument from the dataframe we generated earlier. We will set the `orthoI`
to 1. This is the number of orthogonal components to use in the analysis. We will
set this to 1 for simplicity. We will store the results in the `scsa` object.

The analysis used the labels to generate the conformational signature by using
orthogonal partial least squares discrimant analysis (OPLS-DA). The results return
a formal `opls` object. This object contains the results of the analysis.

Note that the results look strange because we have not run the algorith for
enough iterations earlier. The results are allowed to contain "Unknown" outcomes.

Note that the algorithm will output a number of warnings - 
the first is that the number of predictive components is set to 1. This is the desire behaviour. The second is that the variance in some residue elements is below machine precision.
This is typical for residues that were not measure in the analysis or where the uptake was 0 (such as Prolines). You do not need to worry about this as the sequence labelling is handled behind the scenes.

```{r,}
scsa <- supervisedCSA(RexDifferentialList = out_lxr_compound_proccessed,
                      quantity = "TRE",
                      states = states,
                      labels = labels,
                      whichlabel = "ABCA1",
                      whichTimepoint = 600,
                      orthoI = 1)

scsa2 <- supervisedCSA(RexDifferentialList = out_lxr_compound_proccessed,
                      quantity = "TRE",
                      states = states,
                      labels = labels,
                      whichlabel = "lipogenic",
                      whichTimepoint = 600,
                      orthoI = 1)

```


We can now look at the results of the analysis. The results are stored in the 
`opls` object. For each analysis, we can plot the loading of the conformational
signature. This is the regions of the protein that are most variable across the
states and are also associated with the outcome. We will plot the loadings of the
outcomes. The predictive component is the component that is associated with the
outcome. The orthogonal component is the component that is not associated with the
outcome. We will plot the loadings for ABCA1 and lipogenicity for both 
predictive and orthogonal components.

Note again the results look strange because we have not run the algorithm for
enough iterations earlier.

```{r,}
plotLoadingSCSA(states.plsda = scsa,
                labels = labels,
                whichLoading = "predictive",
                whichlabel = "ABCA1")

plotLoadingSCSA(states.plsda = scsa,
                labels = labels,
                whichLoading = "orthogonal",
                whichlabel = "ABCA1")

plotLoadingSCSA(states.plsda = scsa2,
                labels = labels,
                whichLoading = "predictive",
                whichlabel = "lipogenic")

plotLoadingSCSA(states.plsda = scsa2,
                labels = labels,
                whichLoading = "orthogonal",
                whichlabel = "lipogenic")
```

You can also plot the scores of the conformational signature. This is the
scores of the samples from the predictive and orthogonal components. We will
plot the scores for ABCA1 and lipogenicity for both predictive and orthogonal
components.

```{r,}
plotSCSA(states.plsda = scsa2,
         labels = labels,
         whichlabel = "lipogenic")
```



# Supervised Conformational Signature Analysis with continuous outcomes

We can also perform the supervised conformational signature analysis with continuous
outcomes. This is similar to the discrete outcomes but the outcome is continuous.
To specify the labels we will generate the labels for the outcomes as before. We
will store the results in a dataframe with the states as the rownames. For proteins
where the outcome is unknown, we will set the outcome to `NA`. In this case,
we will use the `ED50` as the outcome. We will log the `ED50` values. You can of
course load this dataframe from something that was made in excel and covert to a
dataframe. We keep these values as numeric rather than factors.

```{r,}
labels$ED50 <- NA
labels[, "ED50"] <- log(c(4.11, NA, NA, 0.956, NA, 9.64, 1.49,
                      5.65, 0.969, 2.10, 11.3, NA, 31.5, 341, 32.2, 17.2, NA))
```

We will now run the supervised conformational signature analysis with continuous
outcomes. We will set the `quantity` to "TRE" as this is the quantity we use to
generate the signature. We will set the `states` to the states we are interested in.
We will set the `whichTimepoint` to 600. This is the timepoint used to generate the
conformational signature. 

To perform the supervised conformational signature analysis with continuous outcomes,
we will use the `supervisedCSA` function. In this case, we will set the `type` to
"continuous" so that it does not perform a discrete analysis. We will set the
`whichlabel` to "ED50" as this is the outcome we are interested in.

```{r,}
scsa3 <- supervisedCSA(RexDifferentialList = out_lxr_compound_proccessed,
                      quantity = "TRE",
                      states = states,
                      labels = labels,
                      whichlabel = "ED50",
                      whichTimepoint = 600,
                      orthoI = 1,
                      type = "continuous")
```

We can now look at the results of the analysis. The results are stored in the
`opls` object. We can plot the loading of the conformational signature. This is
the regions of the protein that are most variable across the states and are also
associated with the outcome. We will also plot the scores of the conformational
signature. This is the scores of the samples on the predictive and orthogonal
components. We will plot the loadings and scores for the `ED50` outcome.

```{r,}

plotSCSA(states.plsda = scsa3,
         labels = labels,
         whichlabel = "ED50",
         type = "continuous") + labs(col = "log ED50")

plotLoadingSCSA(states.plsda = scsa3,
                labels = labels,
                whichLoading = "predictive",
                whichlabel = "ED50")

plotLoadingSCSA(states.plsda = scsa3,
                labels = labels,
                whichLoading = "orthogonal",
                whichlabel = "ED50")
```

# Advanced analysis

Until now we havent used the uncertainty in the analysis. We can use the
uncertainty in the analysis to generate the conformational signature. We sample
from the distribution of the parameters to generate the conformational signature.
We will use the `sampleTREuncertainty` function. This function will take the outputs
of the analysis and sample from the distribution of the parameters to 
quantify the uncertainty in the conformational signature. The important 
argument below is the `num_montecarlo` which is the number of samples to take
from the distribution. We will set this to 5000. The ``whichSamples`` argument
is the samples to use from the orginal analysis. We will use the first 50 samples
- note that since the results have been processed the samples are actually the
first 50 samples of the *processed* results. 

We will not run the code here but pre-load the results from the package.

```{r, eval=FALSE}
TRE_dist <- sampleTREuncertainty(HdxData = LXRalpha_processed,
                     RexParamsList = rex_lxr,
                     quantity = "TRE",
                     states = states,
                     whichChain = 1,
                     whichSamples = seq.int(1, 50),
                     whichTimepoint = 600,
                     num_montecarlo = 5000)

```

We will now load the results from the package. The results are stored in the
`TRE_dist` object.

```{r}
data("TRE_dist")
```

To plot the results we will use the `plotTREuncertaintyLoadings` function. This function
will take the output of the `sampleTREuncertainty` function and plot the results.
We also need to pass the `pca_states` object from the unsupervised analysis. We will
plot the results. This analysis reports the distribution of the loadings.

```{r,}
gg <- plotTREuncertaintyLoadings(df_all = TRE_dist,
                                 pca_states = ucsa$pca_states,
                                 states = states)
gg

```
To visualise the results we can use the `plotTREuncertainty` function. This function
will take the output of the `sampleTREuncertainty` function and plot the results.
We also need to pass the `pca_states` object from the unsupervised analysis. 
This plots the unsupervised analysis with the uncertainty as contours in the pca
coordinates.

Again the results suffer because of the lack of iterations in the analysis.


```{r,}
plotTREuncertainty(df_all = TRE_dist,
                   pca_states = ucsa$pca_states,
                   states = states)

```


You may be interested in plotting the results of the analys on the structure.
This is the same set-up as before you simply need to make sure that you align
the residue numbers correctly

```{r,}
# Using the predictive component of the supervised analysis as an example
library(NGLVieweR)
library(dplyr)

v <- matrix(t(scsa@loadingMN), nrow = 1)

# set colnames carefully to residue number by first remove the x from the rownames
# and then coverting to numeric
colnames(v) <- as.numeric(gsub("x", "", rownames(scsa@loadingMN)))
pdb_filepath <- system.file("extdata", "2acl.pdb", mustWork = TRUE, 
                            package = "RexMS")

mycolor_parameters <- hdx_to_pdb_colours(dataset = v,
                                         pdb = pdb_filepath,
                                         cmap_name = "CSA", scale_limits = c(-0.2,0))

graphics <- NGLVieweR(pdb_filepath) %>%
  stageParameters(backgroundColor = "white", zoomSpeed = 1) %>%
  addRepresentation("ribbon") %>%   setQuality("high") %>% 
  addRepresentation("ribbon", param = list(color=mycolor_parameters,
                                           backgroundColor="white"))

graphics



```



To extract the information from the uncertainty quantification analysis: 


```{r,}
# Using the lower quantile as an example
loadingquants <- gg$data

v <- matrix(as.numeric(loadingquants[loadingquants$level == 0.025,
                                     "quantile", drop = FALSE][[1]]), nrow = 1)
colnames(v) <- seq.int(ncol(v))
pdb_filepath <- system.file("extdata", "2acl.pdb", mustWork = TRUE, 
                            package = "RexMS")


mycolor_parameters <- hdx_to_pdb_colours(dataset = v,
                                         pdb = pdb_filepath,
                                         cmap_name = "CSA")

graphics <- NGLVieweR(pdb_filepath) %>%
  stageParameters(backgroundColor = "white", zoomSpeed = 1) %>%
  addRepresentation("ribbon") %>%   setQuality("high") %>% 
  addRepresentation("ribbon", param = list(color=mycolor_parameters,
                                           backgroundColor="white"))

graphics

```


`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```






