---
title: "Differential HDX-MS with RexMS"
author: 
  - name: Oliver Crook
    affiliation:
    - University of Oxford
    email: ococrook@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('RexMS')`"
vignette: >
  %\VignetteIndexEntry{Differential HDX-MS with RexMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    RexMS = citation("RexMS")[1]
)
```


# Introduction

This vignette demonstrates how to use the `RexMS` package to perform 
differential HDX-MS analysis. We recommend that you start with the introduction
to RexMS vignette first so that you cover the basics. We skip some of those
initial steps here to focus on the aspects important for differential analysis.

# Loading data

We will load the data from the `RexMS` package. This data is a subset of the 
data used in the original publication and focuses on `BRD4` apo and in complex
with I-BET151 both datasets are available as part of the package. Here, we 
are interested in the changes in HD exchange as a results of ligand binding.
To speed up the analysis we filter the data to only include the first 100
residues and again calculate a number of valuable quantities.


```{r load_data}
require(RexMS)
require(dplyr)

data("BRD4_apo")
data("BRD4_ibet")

BRD4_apo <- BRD4_apo %>% filter(End < 100)
BRD4_ibet <- BRD4_ibet %>% filter(End < 100)
numTimepoints <- length(unique(BRD4_apo$Exposure))
Timepoints <- unique(BRD4_apo$Exposure)
numPeptides <- length(unique(BRD4_apo$Sequence))
```

We actually only need to run `rex` explicitly on the condition we are primarily
interested in e.g. 'ligand' vs 'apo', we only need to model the first 
dataset initially. If you swap the analysis around and start with `apo` first
your results will be inverted (e.g. report protection instead of deprotection)
as a consequence of the data. If you need to separate out conditions from
one `csv` file it maybe useful to use `filter(State == "my protein state")`,
where `"my protein state"` would be replace by the state you wishes to extract
from the data. The syntax is similar to the filter line used to extract
the first 100 residues. 


The following function will run the `rex` function on the data in the
ligand bound state and then `RexProcess` will populate the object as in the 
original analysis. Again we would recommend running this function for a much
large number of iterations.

```{r, cache = TRUE}
set.seed(1)
rex_ibet <- rex(HdxData = DataFrame(BRD4_ibet),
                numIter = 100,
                R = max(BRD4_ibet$End), 
                numtimepoints = numTimepoints,
                timepoints = Timepoints,
                density = "laplace",
                seed = 1L,
                tCoef = c(0, rep(1, 5)),
                BPPARAM = SerialParam())

rex_ibet <- RexProcess(HdxData = DataFrame(BRD4_ibet), 
                        params = rex_ibet,
                        range = 50:100,
                        thin = 1,
                        whichChains = c(1,2))

```

We recommend checking the results as you would in the case of a single-protein
analysis but to simplify the analysis we will skip this step here. We plot the
sigmas and log-likelihoods for the ligand bound state below.

```{r plot_results}
library(ggplot2)
plotSigma(rex_ibet)

# zoom in on results a bit but adding the lim arguements
plotLogLikelihoods(rex_ibet) + xlim(c(25,100)) + ylim(c(-500,-250))

```
To obtain differential results as compared to the apo (or some other state)
use the following function carefully. In the `params` argument you want
to pass the object that was created by the `rex` function for the ibet state. 
In the `HdxData` argument you want to pass the data for the apo state (or the
state you wish to compare to). The `whichChain` argument is used to specify
which chain you are interested in using to calculate the differential results.
Chain 2 had a higher log-likelihood in the ligand bound state, so we will use
this chain to calculate the differential results. 

```{r,}
rex_diff <- processDifferential(params = rex_ibet, 
                                HdxData = DataFrame(BRD4_apo),
                                whichChain = c(2))
```

This results in an object of class `RexDifferential` which has five slots:

1. Rex.predictionError - the prediction error for the differential analysis at 
the level of peptides
2. Rex.estimates - the estimates for the differential analysis at the level of
residues
3. Rex.probs - the probabilities for the differential analysis at the level of
residues
4. Rex.eFDR - the estimated false discovery rate for the differential analysis
at various probability thresholds.

We can access these slots as follows and inspect the object

```{r,}
rex_diff
Rex.predictionError(rex_diff)
Rex.estimates(rex_diff)
Rex.probs(rex_diff)
Rex.eFDR(rex_diff)

```



We can plot the results for the differential analysis as follows. The following
will produce a volcano plot of the results. The `nrow` argument specifies the
number of row to facet over. The `quantity` argument specifies the quantity
to plot on the y-axis. The options are `TRE`, or `ARE`. TRE account for the
redundancy at the residue and so is the recommended option. As we can 
see from the plot it is very simple to observe the generaly trend toward
protection/stablisation as the small molecules binds to the protien it will
reduce its flexibility.

```{r plot_diff_results, fig.height=8}
gg1 <- plotVolcano(diff_params = rex_diff,
                   nrow = 5,
                   quantity = "TRE")
gg1
```
For the ARE:

```{r plot_diff_results2, fig.height=8}

gg2 <- plotVolcano(diff_params = rex_diff,
                   nrow = 5,
                   quantity = "ARE")
gg2
```
To add in the spatial context we suggest a line/butterfly plot:

```{r plot_diff_results3, fig.height=8}
gg3 <- plotButterfly(diff_params = rex_diff,
                     nrow = 5,
                     quantity = "TRE")

gg3
gg3 <- plotButterfly(diff_params = rex_diff,
                     nrow = 5,
                     quantity = "signedARE")
gg3

```
If you wish to average over time then we suggest using the following function.
This will average the results over time and produce a line plot. This is usually
a good way to see that some effects are time dependent and not supported at 
all timepoints, whilst for some residues the effect is consistent across all
timepoints.

```{r,}
gg4 <- plotTimeAveragedButterfly(diff_params = rex_diff)
gg4
```
You may also wish to interpret the probabilities carefully
and convert them to an estimated of the expected false discovery rate - i.e. if I picked a probability threshold 
what is the predicted number of false claims I would make. This is stored in the `Rex.eFDR` slot. This will produce
a table of the estimated FDR at various probability thresholds. For example, for a probability threshold of 0.99. 
The eFDR for 15 and 60 seconds is not defined (there are
no values at this threshold for these timepoints). The eFDR for 600 second sis 0.00148 or 0.148%, which is quite low. This
indicates that if you selected a probability threshold of 
0.99 you would expect to make 0.148% false claims from claiming a residue was differentially protected/stabilised when it was not.

```{r,}
Rex.eFDR(rex_diff)
```






`RexMS` supports transferring these results to protien structures using the
NGLVieweR package. This is a very powerful tool for visualising the results
in the context of the protein structure. The following code will produce a
plot of the protein structure with the residues coloured by the differential
(TRE) results at 15 seconds. The `pdb_filepath` argument specifies the path
to the pdb file. The `cmap_name` argument specifies the name of the colour
map to use. The `v` argument specifies the values to use for the colouring.
You must provide a local `pdb_filepath` to use this function it will not
extract the pdb file from the internet. 


```{r,}
library(NGLVieweR)
v <- matrix(Rex.estimates(rex_diff)$TRE_15, nrow = 1)
colnames(v) <- seq.int(ncol(v))
v2 <- v[,seq.int(1, 99), drop = FALSE]
colnames(v2) <- seq.int(ncol(v2)) 

# This complex line is used to get the path to the pdb file from the RexMS package
# yours will probably look like
# pdb_filepath <- "path/to/your/pdb/file.pdb"
pdb_filepath <- system.file("extdata", "test_BRD4.pdb", mustWork = TRUE, package = "RexMS")

mycolor_parameters <- hdx_to_pdb_colours(v2,
                                         pdb = pdb_filepath,
                                         cmap_name = "ProtDeprot")
view_structure(pdb_filepath = pdb_filepath,
               color_parameters = mycolor_parameters)
```
You can change the representation if you wish:

```{r,}
view_structure(pdb_filepath = pdb_filepath,
               color_parameters = mycolor_parameters,
               representation = "spacefill")
```
You can provide any values you would like to colour the structure by. For example
using the time-averaged probabilities.


```{r,}
v <- matrix(Rex.probs(rex_diff)$totalprobs, nrow = 1)
colnames(v) <- seq.int(ncol(v))
v2 <- v[,seq.int(1, 99), drop = FALSE]
colnames(v2) <- seq.int(ncol(v2)) 

mycolor_parameters <- hdx_to_pdb_colours(v2,
                                         pdb = pdb_filepath,
                                         cmap_name = "ProtDeprot")
view_structure(pdb_filepath = pdb_filepath,
               color_parameters = mycolor_parameters)


```

You can also use the `viridis` colour map if you wish and plot random colours.
The things to remember carefully are match the residue numbering to your pdb
file. If the residue numberings is not the same as the pdb file you will get
incorrect results. This is typical when expressing a construct of a protein or
the structure is only a small part of the protein.

```{r,}
library(viridis)
# random colours
v <- matrix(rnorm(100), nrow = 1)
colnames(v) <- seq.int(ncol(v))
v2 <- v[,seq.int(1, 99), drop = FALSE]
colnames(v2) <- seq.int(ncol(v2)) 

mycolor_parameters <- hdx_to_pdb_colours(v2,
                                         pdb = pdb_filepath,
                                         cmap_name = "viridis")
view_structure(pdb_filepath = pdb_filepath,
               color_parameters = mycolor_parameters)


```

This is the end of this tutorial if you have many states and perhaps even
downstream functional outcomes you may wish to visit the next vignette
which covers the ideas of a conformational signature analysis (CSA) where
you can use unsupervised and supervised learning to characterise the states.



# Reproducibility

The `r Biocpkg("RexMS")` package `r Citep(bib[["RexMS"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```








`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and
`r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```