---
title: "Introduction to RexMS"
author: 
  - name: Oliver Crook
    affiliation:
    - University of Oxford
    email: ococrook@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('RexMS')`"
vignette: >
  %\VignetteIndexEntry{Introduction to RexMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    RexMS = citation("RexMS")[1]
)
```

# Basics

## Install `RexMS`

`R` is an open-source statistical environment which can be easily modified
to enhance its functionality via packages. `r Biocpkg("RexMS")` is an `R` 
package available via the [Bioconductor](http://bioconductor.org) repository
for packages. `R` can be installed on any operating
system from [CRAN](https://cran.r-project.org/) after which you
can install `r Biocpkg("RexMS")` by using the following commands in
your `R` session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("RexMS")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

or from installed from github

```{r "install2", eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

remotes::install_github("ococrook/RexMS")
```

## Required knowledge

`r Biocpkg("RexMS")` is based on many other packages 
and in particular in those that have 
implemented the infrastructure needed for dealing with protein structure. 
That is, packages like `r Biocpkg("Bio3D")`.

If you are asking yourself the question "Where do I start using Bioconductor?"
you might be interested in 
[this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

We expect you to have a working understanding of HDX-MS and its application.
Verifying your data is important as `r Biocpkg("RexMS")` won't fix problem with
data acquisition or poor experiment design. If you run into problem using
`r Biocpkg("RexMS")` first check syntex of your command but also the data.

## Asking for help

As package developers, we try to explain clearly how to use our packages
and in which order to use the functions.
But `R` and `Bioconductor` have a steep learning curve so it is critical
to learn where to ask for help. The blog post quoted above mentions some
but we would like to highlight the
[Bioconductor support site](https://support.bioconductor.org/) as the main
resource for getting help: remember to use the `RexMS` tag and
check [the older posts](https://support.bioconductor.org/tag/RexMS/). Other
alternatives are available such as creating GitHub issues and tweeting. However,
please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is
particularly critical that you provide a small reproducible example
and your session information so package developers can track down 
the source of the error.

## Citing `RexMS`

We hope that `r Biocpkg("RexMS")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("RexMS")
```

# Quick start to using `RexMS`

```{r "start", message=FALSE}
library("RexMS")
```


Here is an example of you can cite your package inside the vignette:

* `r Biocpkg("RexMS")` `r Citep(bib[["RexMS"]])`

Let's start with a simple use cases for `r Biocpkg("RexMS")`. We have a protein
with peptide-level HDX-MS data and we wish to gain more granular information
about the HDX pattern at the level of individual residues. We can use RexMS to
perform this analysis. Let's first load some simple HDX-MS data from the 
package.

```{r "simple"}
library(dplyr)

data("BRD4_apo")
BRD4_apo
```

The data is a data frame with the following columns:

```{r,}
colnames(BRD4_apo)
```

The data must contain the following columns: 

"State",
"Sequence",
"Start",
"End",
"MaxUptake",
"Exposure",
"replicate",
"Uptake"

If you only have 1 replicate you can simply set `1` for this column. 

The function `cleanHDX` can be used to clean the data and remove any missing values
but also check that all the required columns exist. The function will output an
error if this simple check fails. The RexMS package needs you to convert
data to a `DataFrame` object.

```{r "clean"}
BRD4_apo <- DataFrame(BRD4_apo)

BRD4_apo <- cleanHDX(res = BRD4_apo, clean = TRUE)

```

If you do not have a `MaxUptake` column you can use the following function. You
may need to use the `rep` function to repeat for time points.

```{r,}
maxUptakes(BRD4_apo)
```


The data is now ready to take a closer look. Be careful at this point you
may have lost a large number of peptides if you had lots of missing values.
At this point it will be useful to check the redundancy of the data. The
following code chunk will perform heatmap of the peptide map. As we can see
from the peptide below, there are two regions that are well convered linked
by a region that is not well covered. This is a common problem in HDX-MS
as the region between the two tandem bromodomains is proline rich and hence
not amenable to standard digestion.

```{r "heatmap"}

C <- coverageHeatmap(res = BRD4_apo)
```

To simplify the analysis, we only examine the first 100 residues of the protein.
This will speed up the results in this vignette. Furthermore, we would recommend
analysis the bromodomains separately.

```{r "subset"}
BRD4_apo <- data.frame(BRD4_apo) %>% filter(End < 100)
```

Rex requires you to switch make to a `DataFrame` if you used the last line

```{r,}
BRD4_apo <- DataFrame(BRD4_apo)
```

The next step is to calculate some simple quantities from the data. The first
is the number of timepoints, the second the numerical values of the timepoints
and the number of peptides in the analysis.

```{r,}
numTimepoints <- length(unique(BRD4_apo$Exposure))
Timepoints <- unique(BRD4_apo$Exposure)
numPeptides <- length(unique(BRD4_apo$Sequence))
```

We already at the point were we can apply Rex to the data. The first step is to
understand the necessary components of the `rex` function. The `rex` function
has a lot of possible arguments which are explain in the documentation. The most 
important are the following

- `HdxData` which is the DataFrame containing the HDX-MS data.
- `numIter` which is the number of iterations to run the algorithm. Typically 
this will be in the thousands - 5000 would be reasonable but is 100 below to 
speed up the algorithm.
- `R` which is the maximum residue number in the protein.
- `numtimepoints` which is the number of timepoints in the data.
- `timepoints` which is the numerical values of the timepoints.
- `seed` which is the seed for the random number generator.
- `tCoef` which is the coefficients for the timepoints. This account for the 
fact that the variance at the 0 timepoint might be 0 (measured without error).
This value is multiplied by the variance across timepoints to scale the variance.
We recommend that the first value is 0 and the rest are 1. However, if you have
a good reason to change this value you can do so.
- `BPPARAM` which is the parallel backend to use. We recommend using the `BiocParallel`
package and the `SerialParam` function. This will allow you to run the algorithm
in parallel on a single machine. If you have multiple cores available you can 
run the algorithm in parrallel to speed up the analysis. To do this use
the `bpparam()` function from the `BiocParallel` package. This will allow you to 
register the default backend. For more advanced approaches you can see the
`BiocParallel` documentation.
- `numChains` which is the number of chains to run in the algorithm. We recommend
running at least 2 chains, which is the default. This will allow you to check
the convergence of the algorithm. If the chains do not converge you can
increase the number of chains and/or the number of iterations.


The less important parameters are the following but still worth knowing about:

- `density` which is the density of the likelihood function, whilst `Gaussian` is
the default we have found better results when using the `laplace` density. These
are the only two options available.
- `R_lower` which is the lower bound of the residues to consider. This is useful
if you have a protein with a signal peptide or a tag that you do not want to
consider in the analysis. Or you have a construct starting at a residue other
than 1.
- `R_upper` which is the upper bound of the residues to consider. This is useful
if you want to contrain the analysis to a specific region of the protein.
- `priors` which is the priors for the analysis. This list is elaborated on later
and we refer to manuscript for more details. We dont recommend changing this
unless you have a good reason to do so.
- `phi` this is the max uptake you allow per residue in the analysis. This might
be the deuterium content of the buffer or the maximum uptake you expect to see.
- `init_param` this is the parameter that used to help initialise the algorithm.
The default is `d`. This is mostly for development purposes and we recommend
leaving this as the default.

Now, you can run the algorithm. The following code chunk will run the algorithm
on the data. The algorithm will take a few minutes to run. The algorithm
may output a warning about a package not being available, you can safely
ignore this warning.

We typically recommend running a short number of iterations first to check 
how long the analysis will take and then running the full analysis. 


```{r "apply Rex", cache=TRUE}
set.seed(1)
rex_test <- rex(HdxData = BRD4_apo,
                numIter = 100,
                R = max(BRD4_apo$End), 
                density = "laplace",
                numtimepoints = numTimepoints,
                timepoints = Timepoints,
                seed = 1L,
                tCoef = c(0, rep(1, numTimepoints - 1)),
                phi = 1,
                BPPARAM = SerialParam())

```
Once the algorithm has finished running you can view the results. First, note
that `rex` produces an object of type `RexParams`. Just printing the object
will give you a summary of the object. 


```{r,}
rex_test
```

The object contains the following slots:

- `interval` which is the interval of the analysis. This is the interval
of the residues that were analysed.
- `chains` which stores all the MCMC iterations of the analysis
- `summary` which is a summary of the chains and the most interesting for 
user. It is currently empty
- `seed` which is the seed used in the analysis.
- `priors` which are the priors used in the analysis.
- `method` which is the method used in the analysis. This will be `ReX` and is
used to track development of the package.

We do not recommend access the slots directly but if you are interested you can
use the `@` . For example:

```{r,}
rex_test@interval
```

The summary slot is currently empty as it has not yet been processed. The 
function `RexProcess` takes a number of arguments and will populate the `summary`
slot. The most important arguments are the following:

- `HdxData` which is the DataFrame containing the HDX-MS data.
- `params` which is the `RexParams` object produced by the `rex` function.
- `range` which is the range of MCMC iterations to consider.
This is useful if you want to remove the burn-in period of the MCMC chain.
- `thin` which is the thinning of the MCMC chain. This is useful if you want
to reduce the size of the MCMC chain. A thin equal to 1 will keep all the 
iterations. A thin equal to 5 will keep every 5th iteration. 
- `whichChains` which is the chains to consider. This is useful if you want to
consider only a subset of the chains. The default is to consider the first two 
chains. You may wish to remove unconverged chains.


```{r,}
rex_test <- RexProcess(HdxData = DataFrame(BRD4_apo), 
                        params = rex_test,
                        range = 50:100,
                        thin = 1,
                        whichChains = c(1,2))

```


The `summary` slot is now populated and you can view the results. We breifly
access the slot using `@` to show the results but again we recommend using
accessor functions.

```{r,}
rex_test@summary

```
A number of functions are available to access the results. The first
is `Rex.globals` which will give you access to the model log likelihood and the
sigma parameter which is a measure of the variance of the data. The lower the 
value of sigma the more resolved tha data is. The model likelihood is a measure
of how well the model fits the data. The higher the value the better the fit.
As we can see from the results below in the first chain the model has fit better
but the results between the two chains a similar. We suggest to examine this
first to check that the chains are behaving similarly.

```{r,}
Rex.globals(rex_test)
```
We next consider the peptide error distribution. This allows us to understand
the modelling error for each peptide. The function `Rex.peptideError` will
give you the error distribution for each peptide i.e. the residual of the fitted
model. This benifit here is two fold. First, making sure that the 
model is performing well and secondly any peptides that may have data that
is inconsistent


```{r,}
Rex.peptideError(rex_test)

```
We can produce a heatmap to quickly examine the residuals. The heatmap show
some important features. First, there is a mixture of `blue` and `red` suggesting
that the model is not biased. If the heatmap is completely red or blue this
indicates some modelling issue. Secondly, the heatmap shows that the errors
are larger for longer peptides. This is expected as the model is more uncertain
about the uptake of longer peptides and the model allows for this is part of
the fitting. The errors will also reduce by running the algorithm for more
iterations. We suggest reporting this figure in manuscript or supplementary 
material.


```{r,}
plotPeptideError(rex_test)

```
It maybe informative view the results in a different way by scaling the
error with peptide length. This will allow you to see if the error is
proportional to the length of the peptide. This is useful as the model
is more uncertain about the uptake of longer peptides. This gives a different
view of the data and suggest that error are mostly 10% of the uptake values but
it some cases close to 20%.

```{r,}
plotPeptideError(HdxData = BRD4_apo,
                 rex_params = rex_test,
                 relative = TRUE)



```
The next step is to examine the residue error distribution. This allows us to
gain a more granular understanding of the error distribution. The function
`Rex.resolution` will give you the error distribution for each residue. This 
completements the global resolution metric sigma and the peptide level error metrics


```{r,}
Rex.resolution(rex_test)
```

To visualise these results, we can use the following function. This will
allow us to quickly examine the results. We suggest not be cautious of 
data outside of 3*sqrt(sigma) as the model is likely struggling or yet to have
converged at this point. As we can see the error distribution is very
high at around residue 68. This corresponds to the peptide which had high
error at peptide-level error distribution. This is expected and the model
is correctly reporting where it is struggling to model that data either because
of lack of convergence or the data is unusual at the point.

```{r, fig.height=8}
# calculate a threshold for the data
sqrt(Rex.globals(rex_test)$sigma)*3

# virualise the residue errors
plotResidueResolution(rex_test, nrow  =5)

```

Further analysis of `rex` outputs is possible. In the diagnostics slots
of the `RexSummary` object, we get an estimator of convergence using parrallel
chains. We recommend running iterations until the estimate is below 1.2 with 
results closer to 1 being generally more stable.

```{r,}
rex_test@summary@diagnostics
```

You can plot both sigma and the model log likelihood as a function of iterations
to diagnose convergence issues. The plots below show that sigma is still decreasing
and the log likelihood still increasing but the chains are tracking each other 
well. We note that all (except heatmaps) are object from `ggplot` and so you
can customise them using gramamatical construction in the same way as
ggplot.

```{r,}

plotSigma(rex_test)
plotLogLikelihoods(rex_test)

# example costumisation
library(ggplot2)
library(RColorBrewer)
plotSigma(rex_test) + scale_colour_manual(values = brewer.pal(n = 3, name = "Pastel1"))

```
The `posteriorEstimates` for the parameters and their 95% confidence can be 
found using the following accessors

```{r,}
posteriorEstimates(rex_test)
Rex.quantiles(rex_test)
colnames(Rex.quantiles(rex_test))
```

<!-Need to write a function to output uptake and uncertainty with plots->


# Reproducibility

The `r Biocpkg("RexMS")` package `r Citep(bib[["RexMS"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("RexMS.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("RexMS.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```








`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and
`r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
